---
title: "Get drug target edges from IUPHAR"
author: "dECMT"
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

require(dplyr)
require(tidyr)

require(neo4r)
require(neo4jshell)

rm(list=ls())

##get today's date
today <- format(Sys.Date(), format = "%d %B %Y")

```

* Although CHEMBL is comprehensive, it does not include MTOR as a target for Everolimus (only FKBP1A)  
* There are likely to be other examples  
* Therefore, supplement drug->target edges from CHEMBL with data from other sources - this case, IUPHAR  
  * See descriptions for downloads here: https://www.guidetopharmacology.org/DATA/file_descriptions.txt  
* No need to create new nodes, as almost all relevant drugs will be represented already in NCI thesaurus (?)  
* We can inject these into the neo4j graph using MERGE, ON CREATE and ON MATCH cypher queries (https://neo4j.com/docs/cypher-manual/current/clauses/merge/#:~:text=true-,Merge%20with%20ON%20CREATE%20and%20ON%20MATCH,-Query )  
* (Note VS-6766 does not appear in NCI thesaurus, CHEMBL or IUPHAR...)  

  
```{r download interactions from IUPHAR }
IUPHAR_interactions_url <- "https://www.guidetopharmacology.org/DATA/interactions.csv"
iuphar_interactions <- read.csv(IUPHAR_interactions_url, skip = 1:1)      ## first row is date refreshed, not a header


## manually download the file from above URL and add to public repo on GitHub where neo4j can access it... 
# IUPHAR_interactions_url <- "https://raw.githubusercontent.com/digital-ECMT/network_data/main/interactions.csv"


## get ligand id mappings
iuphar_ligand_id_mappings <- read.csv("https://www.guidetopharmacology.org/DATA/ligand_id_mapping.csv", skip = 1:1)      ## first row is date refreshed, not a header



```

```{r process iuphar interactions}

iuphar_processed <- dplyr::select(iuphar_interactions, "drug_name"=Ligand, Ligand.ID, Target.UniProt.ID, "action_type"="Type", Selectivity, Primary.Target)

## join to molecule ChEMBL IDs 
iuphar_processed <- merge(x=dplyr::select(iuphar_ligand_id_mappings, ChEMBl.ID, Ligand.id), by.x = "Ligand.id", 
                          y=iuphar_processed, by.y = "Ligand.ID")

## harmonise action type values with those loaded into neo4j... 
action_types <- data.frame("parent_action_type"=character(0), 
                           "action_type"=character(0)) %>%
  add_row(parent_action_type = "NEGATIVE MODULATOR", action_type = "Antagonist") %>%
  add_row(parent_action_type = "NEGATIVE MODULATOR", action_type = "Inhibitor") %>%
  add_row(parent_action_type = "NEGATIVE MODULATOR", action_type = "Gating inhibitor") %>%
  add_row(parent_action_type = "NEGATIVE MODULATOR", action_type = "Channel blocker") %>%
  add_row(parent_action_type = "POSITIVE MODULATOR", action_type = "Agonist") %>%
  add_row(parent_action_type = "POSITIVE MODULATOR", action_type = "Activator") 

## join parent_action_type values
iuphar_processed <- merge(x=iuphar_processed, by.x = "action_type", 
                          y=action_types, by.y = "action_type")

## select and reorder columns
iuphar_processed <- unique(dplyr::select(iuphar_processed, "drug_chembl_id"="ChEMBl.ID", "target_uniprot_id"="Target.UniProt.ID", action_type, parent_action_type, "disease_efficacy"="Primary.Target"))

iuphar_processed$disease_efficacy <- as.logical(iuphar_processed$disease_efficacy)

## add extra columns for consistency with other drug-target edges in neo4j
iuphar_processed$LINK_FUNCTION <- case_when(
    iuphar_processed$parent_action_type == "POSITIVE MODULATOR" ~ 1,
    iuphar_processed$parent_action_type == "NEGATIVE MODULATOR" ~ -1,
    TRUE ~ 0
)
# indicate source
iuphar_processed$source <- "IUPHAR"
# indicate refresh date
iuphar_processed$refresh_date <- today


## drop rows where there is no chembl id
iuphar_processed <- iuphar_processed[iuphar_processed$drug_chembl_id != "", ]
## drop rows where there is no target uniprot id
iuphar_processed <- iuphar_processed[iuphar_processed$target_uniprot_id != "", ]

## some entries have >1 uniprot id, pipe-delimited
# split and unnest
iuphar_processed$target_uniprot_id <- strsplit(iuphar_processed$target_uniprot_id, split = "\\|")
iuphar_processed <- tidyr::unnest(data = iuphar_processed, target_uniprot_id)
iuphar_processed <- as.data.frame(iuphar_processed)
```

```{r write iuphar_processed to csv}

write.csv(x=iuphar_processed, file = "iuphar_processed.csv", row.names = FALSE)
```

### **Upload CSV files to neo4j server**  
    
```{r load credentials from JSON file}
## load credentials data from JSON file
neo4j_credentials <- jsonlite::fromJSON(txt = "neo4j_credentials.json")

neo4j_folder_path <- neo4j_credentials$neo4j_folder_path
neo4j_home_directory_address <- neo4j_credentials$neo4j_home_directory_address
neo4j_base_url <- neo4j_credentials$neo4j_base_url
neo4j_username <- neo4j_credentials$neo4j_username
neo4j_password <- neo4j_credentials$neo4j_password

```
  
* Upload csv file over SSH to directory where they can be ingested into neo4j database  


```{r upload iuphar_processed csv file to remote location NOT USED}

# ## copy csv files to home directory on remote maching hosting neo4j database... 
# # system("scp kegg_edges.csv decmt_por@20.90.21.60:")
# # ## sudo copy that file from home directory to correct folder for loading into neo4j... 
# # system('ssh decmt_por@20.90.21.60 "sudo cp kegg_edges.csv /root/docker/neo4j/neo4jr/import/Network_reasoning/"')
# 
# csv_filenames <- c("kegg_edges.csv", "kegg_nodes.csv", "simple_variant_nodes.csv", "simple_variant_edges.csv", "CIVIC_drug_nodes.csv", "CIVIC_drug_edges.csv")
# 
# for(i in 1:length(csv_filenames)) {
#   filename <- csv_filenames[i]
#   ## copy csv file to home directory on remote maching hosting neo4j database...
#   scp_string1 <- paste0("scp ", filename, " decmt_por@20.90.21.60:")
#   system(scp_string1)
#   ## sudo copy that file from home directory to correct folder for loading into neo4j...
#   scp_string2 <- paste0('ssh decmt_por@20.90.21.60 "sudo cp ', filename,' /root/docker/neo4j/neo4jr/import/Network_reasoning/"')
#   system(scp_string2)
# }
# 
# filename <- "iuphar_processed.csv"
# ## copy csv file to home directory on remote machine hosting neo4j database...
# scp_string1 <- paste0("scp ", filename, " decmt_por@20.90.21.60:")
# system(scp_string1)
# ## sudo copy that file from home directory to correct folder for loading into neo4j...
# scp_string2 <- paste0('ssh decmt_por@20.90.21.60 "sudo cp ', filename,' /root/docker/neo4j/neo4jr/import/Network_reasoning/"')
# system(scp_string2)
```
  
```{r upload csv files to remote location}

## copy csv files to home directory on remote machine hosting neo4j database... 

csv_filenames <- c("iuphar_processed.csv")

for(i in 1:length(csv_filenames)) {
  filename <- csv_filenames[i]
  ## copy csv file to home directory on remote maching hosting neo4j database...
  scp_string1 <- paste0("scp ", 
                       filename, 
                       " ",
                       neo4j_home_directory_address, ":")
  system(scp_string1)
  ## sudo copy that file from home directory to correct folder for loading into neo4j...
  scp_string2 <- paste('ssh', 
      neo4j_home_directory_address, 
      '"sudo cp', 
      filename,
      neo4j_folder_path,
      '"',
      sep = " ")
  system(scp_string2)
}

```

### **Connect to remote DB**  
  
```{r connection credentials}
# base_url <- "bolt+s://neo4jr.digitalecmt.com:17690"
base_url <- "bolt://neo4jr.digitalecmt.com:17690"

username <-  "neo4j"
password <- "marina-sparta-finland-food-precise-5347"

test_query <- "MATCH(g:gene)
WHERE g.entryACCESSION = 'kegg_3845'
RETURN g"
```

```{r run test cypher query using neo4jshell}
## REQUIRES INSTALLATION OF CYPHER SHELL (in this case, in working directory)
shell_path <- paste(getwd(), "cypher-shell", "cypher-shell.bat", sep = "/")

## run a test query
neo4j_query(
    con = list(address = base_url, uid = username, pwd = password),
    qry = test_query,
    shell_path = shell_path,
    database = "neo4j",
)

```


```{r upload to neo4j}
## look for each pair of drug and target nodes... 

## if a relationship doesn't already exist, create one with the defined attributes... 
# MATCH (d:drug), (t:gene)
# WHERE 'CHEMBL3586573' IN d.chembl_id AND 'P42345' IN t.UNIPROT
# MERGE (d)-[r:`NEGATIVE MODULATOR`]->(t)
#     ON CREATE SET r.source = 'IUPHAR test', r.action_type = 'INHIBITOR'

# upload_iuphar_edges_query <- "MATCH (d:drug), (t:gene)
# WHERE 'CHEMBL3586573' IN d.chembl_id AND 'P42345' IN t.UNIPROT
# MERGE (d)-[r:`NEGATIVE MODULATOR`]->(t)
#     ON CREATE SET r.source = 'IUPHAR test', r.action_type = 'INHIBITOR'"



path_to_edges <- "file:///Network_reasoning/iuphar_processed.csv"
## note we are using apoc.MERGE.relationship ... 

                                 
  

## if edges don't exist, create them... 
cypher_MergeIUPHAREdges <- paste0("LOAD CSV WITH HEADERS FROM '",path_to_edges,"' AS edges
                                 MATCH (n1:drug) WHERE edges.drug_chembl_id IN n1.chembl_id
                                 MATCH (n2:gene) WHERE edges.target_uniprot_id IN n2.UNIPROT
                                 CALL {WITH n1, n2, edges
                                 CALL apoc.merge.relationship(n1,edges.parent_action_type,
                                {LINK_FUNCTION:edges.LINK_FUNCTION}, 
                                {action_type:edges.action_type, disease_efficacy:edges.disease_efficacy, source:edges.source, LINK_FUNCTION:edges.LINK_FUNCTION, refresh_date:edges.refresh_date}, 
                                n2, 
                                {action_type:edges.action_type, disease_efficacy:edges.disease_efficacy, source:edges.source, LINK_FUNCTION:edges.LINK_FUNCTION, refresh_date:edges.refresh_date}
                                )
YIELD rel
RETURN NULL}
RETURN null")

## only issue with this is, if the edge already exists (e.g. from NCI thesaurus, the source property of the edge will be set to "IUPHAR", not ["NCIt", "IUPHAR"])
# potentially used coalesce function, as described here: https://groups.google.com/g/neo4j/c/bY5GXWYpM1k#:~:text=%EE%97%93-,Russell%20Duhon,-unread%2C 
                              

neo4j_query(
    con = list(address = base_url, uid = username, pwd = password),
    qry = cypher_MergeIUPHAREdges,
    shell_path = shell_path,
    database = "neo4j",
)


## test
neo4j_query(
    con = list(address = base_url, uid = username, pwd = password),
    qry = "MATCH path = (d:drug)-[r]->(t:gene)
WHERE r.source = 'IUPHAR'
    RETURN path LIMIT 1",
    shell_path = shell_path,
    database = "neo4j",
)

```


`r knitr::knit_exit()`  


  


