---
title: "getKEGGNetworks"
author: "dECMT"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
##NOTE: the following packages are required to run this script, but should be installed (e.g. using code snippets below) before runnning the script, NOT as part of the script itself
# options(repos = "http://cran.us.r-project.org")
# install.packages("BiocManager")
# BiocManager::install("AnnotationDbi")
# BiocManager::install("org.Hs.eg.db")
# BiocManager::install("KEGGREST")
# BiocManager::install("KEGGlincs")
# BiocManager::install("hgu133a.db")
# BiocManager::install("drugTargetInteractions")
require(KEGGlincs)        ## GPL-3
require(KEGGgraph)        ## GPL >= 2
require(KEGGREST)         ## Artistic 2.0
require(dplyr)
require(tidyr)
require(formattable)
require(kableExtra)
require(stringr)
require(neo4r)
require(neo4jshell)
require(httr)
require(jsonlite)
require(reticulate)
require(purrr)
require(corpus)
require(quanteda)
require(org.Hs.eg.db)

## clean up
rm(list=ls())
```

## **Objective**  
  
* Get gene/protein, compound relationships for cancer pathways from KEGG  
  * Save node and egde lists as csv files  
  * Update remote neo4j database  
* Can be run periodically to update graph (without deleting whole graph first...)  

    
## **KEGG networks**  
  
### **Download pathways from KEGG**  
  
```{r download KGML files from KEGG}

## PTEN-|PIP3 is not encoded within hsa05200 Pathways in Cancer (there, it is encoded as "missing interaction")
## PTEN -| PIP3 is encoded within hsa04151 PI3K-AKT signalling pathway, however

## hsa04151 is embedded within hsa05200

## we can use the function parseKGMLexpandMaps() to parse hsa05200, download KGML files for embedded pathways from KEGG FTP website, parse them individually, and merge all the children pathway graphs with the parental pathway into one graph object. The graph is returned as the function value. 
## first, we need to manually download and save KGML file
# embedded_graph <- parseKGMLexpandMaps(file="hsa05200.xml", downloadmethod = "curl", genesOnly = FALSE)
## however, the resulting graph object contains KEGG IDs as identifiers for nodes, and does not contain details for edges (e.g. inhibition etc)

## alternatively, we can get the names of entries of TYPE "map" from pathway_mappings object, then use KEGGPATHID2NAME mapping in KEGGgraph to get IDs for embedded pathways
## however, it is unclear how to implement this (?!)

## for now, we will do this manually - based on the parseKGMLexpandMaps call above, we can see that the following pathways are embedded within pathways in cancer: 
# "hsa04520", "hsa04310", "hsa04512", "hsa04060", "hsa04630", "hsa04010", "hsa04150", "hsa04210", "hsa04115", "hsa04110", "hsa04370", "hsa03320", "hsa04350", "hsa04510", "hsa04151", "hsa04024", "hsa04330", "hsa04340", "hsa04915", "hsa04066", "hsa04020"

cancer_pathways <- c("hsa05200", "hsa05210", "hsa05212", "hsa05225", "hsa05226", "hsa05214", "hsa05216", "hsa05221", "hsa05220", "hsa05217", "hsa05218", "hsa05211", "hsa05219", "hsa05215", "hsa05213", "hsa05224", "hsa05222", "hsa05223")

embedded_pathways <- c("hsa04520", "hsa04310", "hsa04512", "hsa04060", "hsa04630", "hsa04010", "hsa04150", "hsa04210", "hsa04115", "hsa04110", "hsa04370", "hsa03320", "hsa04350", "hsa04510", "hsa04151", "hsa04024", "hsa04330", "hsa04340", "hsa04915", "hsa04066", "hsa04020")

combined_pathways <- unique(c(cancer_pathways, embedded_pathways))

## initialise empty dataframes that will hold nodes and edges
combined_nodelist <- NULL
combined_edgelist <- NULL

## and an empty dataframe that will hold pathway titles
pathway_titles <- data.frame(kegg_pathway_id = character(0), pathway_title = character(0))

## loop through set of pathway IDs and download graph data
for(i in 1:length(combined_pathways)) {
  pathway_id <- combined_pathways[i]
  # print(pathway_id)
  ## download KGML file 
  pathway_KGML <- tryCatch({get_KGML(pathway_id)}, 
                           error = function(cond) {return(NULL)})
  if (is.null(pathway_KGML)) next   ## break this loop and go to next loop, i+1
  ## get pathway title
  pathwayTitle <- getTitle(pathway_KGML)
  # print(pathwayTitle)
  
  ## add rows
  pathway_titles <- pathway_titles %>%
    dplyr::add_row(kegg_pathway_id = pathway_id, pathway_title = pathwayTitle)
  ## expand mappings, convert KEGG IDs to Entrez Symbols
  pathway_mappings <- expand_KEGG_mappings(pathway_KGML)
  pathway_edges <- expand_KEGG_edges(pathway_KGML, pathway_mappings)

  ## add "kegg_" as a prefix to accession numbers
  pathway_edges$entry1accession <- paste0("kegg_", pathway_edges$entry1accession)
  pathway_edges$entry2accession <- paste0("kegg_", pathway_edges$entry2accession)

  ## create a simple edgelist
  edgelist <- unique(dplyr::select(pathway_edges, "from"="entry1accession", "to"="entry2accession", "relationship_detail"="specific_subtype"))
  # specify source
  edgelist$source <- "KEGG"
  
  # create simple node list
  nodelist <- unique(dplyr::select(pathway_edges, "entryACCESSION"="entry1accession", "entrez_symbol"="entry1symbol", "entryTYPE"="entry1type"))
  nodelist2 <- unique(dplyr::select(pathway_edges, "entryACCESSION"="entry2accession", "entrez_symbol"="entry2symbol", "entryTYPE"="entry2type"))
  nodelist <- unique(rbind(nodelist, nodelist2))
  rm(nodelist2)
  # specify source
  nodelist$source <- "KEGG"
  ## specify pathway
  nodelist$pathway <- pathwayTitle
  
  ## bind onto combined lists
  combined_nodelist <- unique(rbind(combined_nodelist, nodelist))
  combined_edgelist <- unique(rbind(combined_edgelist, edgelist))
}  
  
## clean up
# rm(pathway_KGML)
# rm(pathway_mappings)
# rm(pathway_edges)
```
  
* KEGG data for the following pathways have been downloaded and saved as a non-redundant set of nodes and edges:  
  * `r formattable(pathway_titles)`  

  
* The following relationship types are included:  
  * `r unique(combined_edgelist$relationship_detail)`  
  
* For simplicity and network reasoning, we will group these as follows:    
  

```{r specify grouping of relationship types}
relationship_groupings <- data.frame(detail = unique(combined_edgelist$relationship_detail), 
                                     relationship = "other")

relationship_groupings$relationship[grep(pattern = "activation|expression", x=relationship_groupings$detail, ignore.case = T)] <- "activates"

relationship_groupings$relationship[grep(pattern = "inhibition|repression", x=relationship_groupings$detail, ignore.case = T)] <- "inhibits"

## add a LINK_FUNCTION attribute
relationship_groupings$LINK_FUNCTION <- 0
relationship_groupings$LINK_FUNCTION[relationship_groupings$relationship == "activates"] <- 1
relationship_groupings$LINK_FUNCTION[relationship_groupings$relationship == "inhibits"] <- -1

formattable(relationship_groupings)

## join to edge list
combined_edgelist <- unique(merge(x=combined_edgelist, by.x = "relationship_detail", all.x = TRUE, y=relationship_groupings, by.y = "detail"))
```
  
* **We will omit all of the following relationships:**  
  * ***detail* = *activation_inhibition* as this is ambiguous**  

```{r omit relationships}
combined_edgelist <- unique(dplyr::filter(combined_edgelist, relationship_detail != "activation_inhibition"))

# combined_edgelist <- unique(dplyr::filter(combined_edgelist, relationship != "other"))

```

* Where there is >1 relationship remaining between two given nodes, these are aggregated in pipe-delimited format  

```{r group combined_edgelist}
## for a given relationship, there may be >1 relationship_detail
combined_edgelist <- combined_edgelist %>%
  group_by(across(-c(relationship_detail))) %>%
  summarise(
    relationship_detail = paste(unique(na.omit(relationship_detail)), collapse = "|") 
  ) %>%
  as.data.frame()


## reorder columns
combined_edgelist <- unique(dplyr::select(combined_edgelist, from, to, relationship, relationship_detail, source, LINK_FUNCTION))
```
  
  
### **Add metadata to node lists**  
   
* Classify genes as tumour suppressors or oncogenes based on Intogen annotations  
  
```{r get intogen annotations}

compendium_url <- "https://www.intogen.org/download?file=IntOGen-Drivers-20200201.zip"

## download, extract the file from this URL and save Compendium_Cancer)Genes.tsv to the working directory...

compendium <- read.table("Compendium_Cancer_Genes.tsv", sep="\t", quote="", fill = TRUE, stringsAsFactors = FALSE, header = TRUE)

## we just need the ROLE column, defined in README file as follows: 
## ROLE: Consensus mode of action of the gene. Derived from intOGen and from literature. LoF (Loss of Function), Act (Activating) or amb (Ambiguous). 
## also this in the web front end: "Predicted ROLE in pancancer analysis (Actviating for oncogens, loss of function for tumour suppressors, and ambiguous when the estimation is in the twilight zone"

## select columns of interest
compendium <- unique(dplyr::select(compendium, SYMBOL, ROLE))

## reword ROLE values
compendium$ROLE <- gsub(pattern = "Act", replacement = "ONCOGENE", x=compendium$ROLE)
compendium$ROLE <- gsub(pattern = "ambiguous", replacement = "AMBIGUOUS", x=compendium$ROLE)
compendium$ROLE <- gsub(pattern = "LoF", replacement = "TUMOUR SUPPRESSOR", x=compendium$ROLE)

# compendium <- rename(compendium, "driver_mechanism"="ROLE")

compendium <- unique(dplyr::filter(compendium, ROLE != "AMBIGUOUS"))


## check which of the 4 genes from test case are included
intersect(c("ARID1A", "BRAF", "KRAS", "PIK3CA"), compendium$SYMBOL)

```

```{r join intogen annotations to nodelist}
## left join to cancerNodes, keeping all nodes, including those (e.g. compounds) that don't map to an OMIM ID
combined_nodelist <- unique(merge(x=combined_nodelist, by.x = "entrez_symbol", all.x = TRUE, y=compendium, by.y = "SYMBOL", all.y=FALSE))

```

   
* OMIM IDs (where applicable) will be included as additional metadata for KEGG nodes.  
  * Where there is >1 OMIM ID for a given node, these are aggregated in pipe-delimited format  
  * Where a node participates in >1 pathway, pathways are aggregated in pipe-delimited format  
    
```{r get OMIM and UniProt IDs for KEGG entities}

## get OMIM IDs for all nodes based on Entrez Symbol
# first, define columns...
# cols <- c("SYMBOL", "GENENAME", "ENSEMBL", "OMIM", "UNIPROT)
cols <- c("SYMBOL", "OMIM", "UNIPROT" )
omimIDs <- as.data.frame(AnnotationDbi::select(org.Hs.eg.db, keys=combined_nodelist$entrez_symbol, columns=cols, keytype="SYMBOL"))

## left join to cancerNodes, keeping all nodes, including those (e.g. compounds) that don't map to an OMIM ID
combined_nodelist <- unique(merge(x=combined_nodelist, by.x = "entrez_symbol", all.x = TRUE, y=omimIDs, by.y = "SYMBOL", all.y=FALSE))

rm(omimIDs)

## note that there may be >1 OMIM ID for each gene... 
# group OMIM IDs into a single row per gene, pipe-delimited
combined_nodelist <- combined_nodelist %>%
  # group_by(entryACCESSION, entrez_symbol) %>%
  group_by(across(-c(OMIM, UNIPROT, pathway))) %>%
  summarise(
    OMIM = paste(unique(na.omit(OMIM)), collapse = "|"), 
    UNIPROT = paste(unique(na.omit(UNIPROT)), collapse = "|"), 
    pathway = paste(unique(na.omit(pathway)), collapse = "|")
  ) %>%
  as.data.frame()

## reorder columns

combined_nodelist <- unique(dplyr::select(combined_nodelist, entryACCESSION, entrez_symbol, entryTYPE, OMIM, UNIPROT, ROLE, pathway, source))

```
  
* We will define additional node attributes as follows:  
  * *activity*, assigned a value of 1  
  * *deviation*, assigned a value of 0  
  * *label*, duplicating *entrez_symbol* value  
  

```{r create activity attribute for nodes}
## if network effects are multiplicative (e.g. AKT2 effect on MTOR multiplies effect of AKT1 on MTOR), define baseline activity as 1 and then multiply accordingly...
combined_nodelist$activity <- 1

## if network effects are additive (e.g. AKT2 effect on MTOR is added to AKT1 effect on MTOR), define baseline deviation as 0, then add/delete accordingly
combined_nodelist$deviation <- 0

combined_nodelist$label <- combined_nodelist$entrez_symbol
```

### **Save to CSV**  
  
* KEGG node and edge lists have been saved to csv files in the following format:  
  
```{r replace empty strings with NA}
## Replace on all columns
combined_nodelist[combined_nodelist == ''] <- NA
combined_edgelist[combined_edgelist == ''] <- NA

```


```{r indicate refresh date}
##get today's date
today <- format(Sys.Date(), format = "%d %B %Y")
combined_nodelist$refresh_date <- today
combined_edgelist$refresh_date <- today

```

  
```{r save edge and node lists to csv}
## write to csv
write.csv(x=combined_nodelist, file = "kegg_nodes.csv", row.names = FALSE)
write.csv(x=combined_edgelist, file = "kegg_edges.csv", row.names = FALSE)

## preview
formattable(head(combined_edgelist))

formattable(head(combined_nodelist))
```

* Collectively, this KEGG network includes **`r nrow(combined_edgelist)` relationships between `r length(unique(combined_nodelist$entrez_symbol))` nodes**  
  
### **Upload CSV files to neo4j server**  
    
```{r load credentials from JSON file}
## load credentials data from JSON file
neo4j_credentials <- jsonlite::fromJSON(txt = "neo4j_credentials.json")

neo4j_folder_path <- neo4j_credentials$neo4j_folder_path
neo4j_home_directory_address <- neo4j_credentials$neo4j_home_directory_address
neo4j_base_url <- neo4j_credentials$neo4j_base_url
neo4j_username <- neo4j_credentials$neo4j_username
neo4j_password <- neo4j_credentials$neo4j_password

```
  
* Upload csv file over SSH to directory where they can be ingested into neo4j database  
  
```{r upload csv files to remote location}

## copy csv files to home directory on remote maching hosting neo4j database... 

csv_filenames <- c("kegg_edges.csv", "kegg_nodes.csv")

for(i in 1:length(csv_filenames)) {
  filename <- csv_filenames[i]
  ## copy csv file to home directory on remote maching hosting neo4j database...
  scp_string1 <- paste("scp", 
                       filename, 
                       neo4j_home_directory_address, 
                       sep = " ")
  system(scp_string1)
  ## sudo copy that file from home directory to correct folder for loading into neo4j...
  scp_string2 <- paste('ssh', 
      neo4j_home_directory_address, 
      '"sudo cp', 
      filename,
      neo4j_folder_path,
      '"',
      sep = " ")
  system(scp_string2)
}

```


### **Connect to remote DB**  
  
```{r connection credentials}
base_url <- neo4j_base_url

username <-  neo4j_username
password <- neo4j_password

test_query <- "MATCH(g:gene)
WHERE g.entryACCESSION = 'kegg_3845'
RETURN g"
```

```{r run test cypher query using neo4jshell}
## REQUIRES INSTALLATION OF CYPHER SHELL (in this case, in working directory)
shell_path <- paste(getwd(), "cypher-shell", "cypher-shell.bat", sep = "/")

## run a test query
neo4j_query(
    con = list(address = base_url, uid = username, pwd = password),
    qry = test_query,
    shell_path = shell_path,
    database = "neo4j",
)

```

### **Merge KEGG nodes into neo4j DB**  
  
```{r load nodelist into neo4j}

path_to_kegg_nodes <- "file:///Network_reasoning/kegg_nodes.csv"

## using apoc.MERGE.node ... 
## look for each kegg node... 
## if it doesn't already exist, create one with the defined attributes... 
## define query
cypher_MergeKeggNodes <- paste0("LOAD CSV WITH HEADERS FROM '", 
                                 path_to_kegg_nodes,
                                 "' AS nodes 
                                 WITH nodes 
        CALL apoc.merge.node([nodes.entryTYPE],
            {label:nodes.label,
            entryACCESSION:nodes.entryACCESSION,
            entrez_symbol:nodes.entrez_symbol,
            role:nodes.ROLE,
            entryTYPE:nodes.entryTYPE,
            source:nodes.source,
            OMIM:split(nodes.OMIM,'|'),
            UNIPROT:split(nodes.UNIPROT,'|'),
            activity:nodes.activity,
            deviation:nodes.deviation
            },
            {refresh_date: nodes.refresh_date, pathway:split(nodes.pathway,'|')},
            {refresh_date: nodes.refresh_date, pathway:split(nodes.pathway,'|')}
            )
        YIELD node
        RETURN null")

                     
## run query
neo4j_query(
    con = list(address = base_url, uid = username, pwd = password),
    qry = cypher_MergeKeggNodes,
    shell_path = shell_path,
    database = "neo4j",
)



```

### **Merge KEGG edges into neo4j DB**  
  
```{r merge kegg edges into neo4j DB}

path_to_kegg_edges <- "file:///Network_reasoning/kegg_edges.csv"

## if edges don't exist, create them... 
cypher_MergeKeggEdges <- paste0("LOAD CSV WITH HEADERS FROM '",path_to_kegg_edges,"' AS edges
                                 MATCH (n1) WHERE edges.from = n1.entryACCESSION
                                 MATCH (n2) WHERE edges.to = n2.entryACCESSION
                                 CALL {WITH n1, n2, edges
                                 CALL apoc.merge.relationship(n1,edges.relationship,
                                {LINK_FUNCTION:edges.LINK_FUNCTION}, 
                                {relationship_detail:edges.relationship_detail, LINK_FUNCTION:edges.LINK_FUNCTION, source:edges.source, refresh_date:edges.refresh_date}, 
                                n2, 
                                {refresh_date:edges.refresh_date}
                                )
YIELD rel
RETURN NULL}
RETURN null")

## run query
neo4j_query(
    con = list(address = base_url, uid = username, pwd = password),
    qry = cypher_MergeKeggEdges,
    shell_path = shell_path,
    database = "neo4j",
)


```

   
`r knitr::knit_exit()`  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
  
## **Variant-gene relationships**  
  
* Clinvar includes detailed information for approx 75 variants for each gene  
* When creating a graph, this means that there are many more variant:gene relationships compared with gene:gene relationships, which creates performance issues (in the absence of optimisation)  
* for proof of concept, a generic type of variant for each gene (e.g. KRAS GoF or PTEN LoF) will be sufficient  
* This information is available from Intogen at: https://www.intogen.org/download  
  
  
```{r load kegg nodes of interest from csv}
## we are only interested in variants that relate to nodes in the graph 

# load csv
# kegg_nodes <- read.csv(file = "kegg_nodes.csv", header = TRUE, stringsAsFactors = FALSE)
## we will map rfom variants to gene based on (Entrez) symbol, so just get the set of gene symbols for the nodes

kegg_node_symbols <- unique(combined_nodelist$entrez_symbol)


## check which of the 4 genes from test case are included
intersect(c("ARID1A", "BRAF", "KRAS", "PIK3CA"), kegg_node_symbols)
```

```{r create simple variant node list}

simple_variant_nodes <- unique(dplyr::select(compendium, "entrez_symbol"="SYMBOL", "driver_mechanism"="ROLE"))

## replace driver_mechanism values with GoF or LoF
simple_variant_nodes$driver_mechanism[simple_variant_nodes$driver_mechanism == "ONCOGENE"] <- "GoF"
simple_variant_nodes$driver_mechanism[simple_variant_nodes$driver_mechanism == "TUMOUR SUPPRESSOR"] <- "LoF"


## add a label that will be used in neo4j
simple_variant_nodes$label <- paste(simple_variant_nodes$entrez_symbol, simple_variant_nodes$driver_mechanism, sep = " ")

## filter out rows where driver_mechanism is unknown
simple_variant_nodes <- unique(dplyr::filter(simple_variant_nodes, driver_mechanism %in% c("GoF", "LoF")))

## check which of the 4 genes from test case are included
intersect(c("ARID1A", "BRAF", "KRAS", "PIK3CA"), simple_variant_nodes$entrez_symbol)
```

```{r create simple variant edge list}

simple_variant_edges <- unique(dplyr::select(simple_variant_nodes, "from"="label", "to"="entrez_symbol", driver_mechanism))

## assign a value for relationship based on whether driver is GoF or LoF... 
simple_variant_edges$relationship[simple_variant_edges$driver_mechanism == "GoF"] <- "activates"
simple_variant_edges$relationship[simple_variant_edges$driver_mechanism == "LoF"] <- "inhibits"


## add LINK_FUNCTION column
simple_variant_edges$LINK_FUNCTION <- 0
simple_variant_edges$LINK_FUNCTION[simple_variant_edges$driver_mechanism == "GoF"] <- 1
simple_variant_edges$LINK_FUNCTION[simple_variant_edges$driver_mechanism == "LoF"] <- -1

```

* Simple node and edge lists have been saved to CSV in the following format:  
  
```{r save simple node and edge lists to csv}
## write to csv
write.csv(x=simple_variant_nodes, file = "variant_nodelist_intogen.csv", row.names = FALSE)

## preview
formattable(head(simple_variant_nodes))


## write to csv
write.csv(x=simple_variant_edges, file = "variant_edgelist_intogen.csv", row.names = FALSE)

## preview
formattable(head(simple_variant_edges))
```
  
  
  
  
  