---
title: "Get network variants"
author: "dECMT"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
##NOTE: the following packages are required to run this script, but should be installed (e.g. using code snippets below) before runnning the script, NOT as part of the script itself
# options(repos = "http://cran.us.r-project.org")
# install.packages("BiocManager")
# BiocManager::install("AnnotationDbi")
# BiocManager::install("org.Hs.eg.db")
# BiocManager::install("KEGGREST")
# BiocManager::install("KEGGlincs")
# BiocManager::install("hgu133a.db")
# BiocManager::install("drugTargetInteractions")
require(KEGGlincs)        ## GPL-3
require(KEGGgraph)        ## GPL >= 2
require(KEGGREST)         ## Artistic 2.0
require(dplyr)
require(tidyr)
require(formattable)
require(kableExtra)
require(stringr)
require(neo4r)
require(neo4jshell)
require(httr)
require(jsonlite)
require(reticulate)
require(purrr)
require(corpus)
require(quanteda)
require(org.Hs.eg.db)

## clean up
rm(list=ls())
```

## **Objective**  
  
* Get variant-gene relationships for cancer drivers from Intogen  
  * Save node and egde lists as csv files  
  * Update remote neo4j database  
* Can be run periodically to update variants in graph (without deleting whole graph first...)  

  
```{r get intogen annotations}

compendium_url <- "https://www.intogen.org/download?file=IntOGen-Drivers-20200201.zip"

## download, extract the file from this URL and save Compendium_Cancer)Genes.tsv to the working directory...

compendium <- read.table("Compendium_Cancer_Genes.tsv", sep="\t", quote="", fill = TRUE, stringsAsFactors = FALSE, header = TRUE)

## we just need the ROLE column, defined in README file as follows: 
## ROLE: Consensus mode of action of the gene. Derived from intOGen and from literature. LoF (Loss of Function), Act (Activating) or amb (Ambiguous). 
## also this in the web front end: "Predicted ROLE in pancancer analysis (Actviating for oncogens, loss of function for tumour suppressors, and ambiguous when the estimation is in the twilight zone"

## select columns of interest
compendium <- unique(dplyr::select(compendium, SYMBOL, ROLE))

## reword ROLE values
compendium$ROLE <- gsub(pattern = "Act", replacement = "ONCOGENE", x=compendium$ROLE)
compendium$ROLE <- gsub(pattern = "ambiguous", replacement = "AMBIGUOUS", x=compendium$ROLE)
compendium$ROLE <- gsub(pattern = "LoF", replacement = "TUMOUR SUPPRESSOR", x=compendium$ROLE)

# compendium <- rename(compendium, "driver_mechanism"="ROLE")

compendium <- unique(dplyr::filter(compendium, ROLE != "AMBIGUOUS"))


## check which of the 4 genes from test case are included
intersect(c("ARID1A", "BRAF", "KRAS", "PIK3CA"), compendium$SYMBOL)

```

```{r create simple variant node list}

simple_variant_nodes <- unique(dplyr::select(compendium, "entrez_symbol"="SYMBOL", "driver_mechanism"="ROLE"))

## replace driver_mechanism values with GoF or LoF
simple_variant_nodes$driver_mechanism[simple_variant_nodes$driver_mechanism == "ONCOGENE"] <- "GoF"
simple_variant_nodes$driver_mechanism[simple_variant_nodes$driver_mechanism == "TUMOUR SUPPRESSOR"] <- "LoF"


## add a label that will be used in neo4j
simple_variant_nodes$label <- paste(simple_variant_nodes$entrez_symbol, simple_variant_nodes$driver_mechanism, sep = " ")

## filter out rows where driver_mechanism is unknown
simple_variant_nodes <- unique(dplyr::filter(simple_variant_nodes, driver_mechanism %in% c("GoF", "LoF")))

simple_variant_nodes$source <- "Intogen"

## check which of the 4 genes from test case are included
intersect(c("ARID1A", "BRAF", "KRAS", "PIK3CA"), simple_variant_nodes$entrez_symbol)
```

```{r create simple variant edge list}

simple_variant_edges <- unique(dplyr::select(simple_variant_nodes, "from"="label", "to"="entrez_symbol", driver_mechanism))

## assign a value for relationship based on whether driver is GoF or LoF... 
simple_variant_edges$relationship[simple_variant_edges$driver_mechanism == "GoF"] <- "activates"
simple_variant_edges$relationship[simple_variant_edges$driver_mechanism == "LoF"] <- "inhibits"


## add LINK_FUNCTION column
simple_variant_edges$LINK_FUNCTION <- 0
simple_variant_edges$LINK_FUNCTION[simple_variant_edges$driver_mechanism == "GoF"] <- 1
simple_variant_edges$LINK_FUNCTION[simple_variant_edges$driver_mechanism == "LoF"] <- -1


simple_variant_edges$source <- "Intogen"
```

```{r indicate refresh date}
##get today's date
today <- format(Sys.Date(), format = "%d %B %Y")
simple_variant_nodes$refresh_date <- today
simple_variant_edges$refresh_date <- today

```

* Simple node and edge lists have been saved to CSV in the following format:  
  
```{r save simple node and edge lists to csv}
## write to csv
write.csv(x=simple_variant_nodes, file = "variant_nodelist_intogen.csv", row.names = FALSE)

## preview
formattable(head(simple_variant_nodes))


## write to csv
write.csv(x=simple_variant_edges, file = "variant_edgelist_intogen.csv", row.names = FALSE)

## preview
formattable(head(simple_variant_edges))
```
  
### **Upload CSV files to neo4j server**  
    
```{r load credentials from JSON file}
## load credentials data from JSON file
neo4j_credentials <- jsonlite::fromJSON(txt = "neo4j_credentials.json")

neo4j_folder_path <- neo4j_credentials$neo4j_folder_path
neo4j_home_directory_address <- neo4j_credentials$neo4j_home_directory_address
neo4j_base_url <- neo4j_credentials$neo4j_base_url
neo4j_username <- neo4j_credentials$neo4j_username
neo4j_password <- neo4j_credentials$neo4j_password

```
  
* Upload csv file over SSH to directory where they can be ingested into neo4j database  
  
```{r upload csv files to remote location}

## copy csv files to home directory on remote machine hosting neo4j database... 

csv_filenames <- c("variant_nodelist_intogen.csv", "variant_edgelist_intogen.csv")

for(i in 1:length(csv_filenames)) {
  filename <- csv_filenames[i]
  ## copy csv file to home directory on remote maching hosting neo4j database...
  scp_string1 <- paste0("scp ", 
                       filename, 
                       " ",
                       neo4j_home_directory_address, ":")
  system(scp_string1)
  ## sudo copy that file from home directory to correct folder for loading into neo4j...
  scp_string2 <- paste('ssh', 
      neo4j_home_directory_address, 
      '"sudo cp', 
      filename,
      neo4j_folder_path,
      '"',
      sep = " ")
  system(scp_string2)
}

```

### **Connect to remote DB**  
  
```{r connection credentials}
base_url <- neo4j_base_url

username <-  neo4j_username
password <- neo4j_password

test_query <- "MATCH(g:gene)
WHERE g.entryACCESSION = 'kegg_3845'
RETURN g"
```

```{r run test cypher query using neo4jshell}
## REQUIRES INSTALLATION OF CYPHER SHELL (in this case, in working directory)
shell_path <- paste(getwd(), "cypher-shell", "cypher-shell.bat", sep = "/")

## run a test query
neo4j_query(
    con = list(address = base_url, uid = username, pwd = password),
    qry = test_query,
    shell_path = shell_path,
    database = "neo4j",
)

```

### **Merge variant nodes into neo4j DB**  
  
```{r load nodelist into neo4j}

path_to_nodes <- "file:///Network_reasoning/variant_nodelist_intogen.csv"

## using apoc.MERGE.node ... 
## look for each kegg node... 
## if it doesn't already exist, create one with the defined attributes... 
## define query
cypher_MergeVariantNodes <- paste0("LOAD CSV WITH HEADERS FROM '", 
                                 path_to_nodes,
                                 "' AS nodes 
                                 WITH nodes 
        CALL apoc.merge.node(['variant'],
            {label:nodes.label,
            entrez_symbol:nodes.entrez_symbol,
            driver_mechanism:nodes.driver_mechanism
            },
            {source:nodes.source, refresh_date: nodes.refresh_date},
            {source:nodes.source, refresh_date: nodes.refresh_date}
            )
        YIELD node
        RETURN null")

                     
## run query
neo4j_query(
    con = list(address = base_url, uid = username, pwd = password),
    qry = cypher_MergeVariantNodes,
    shell_path = shell_path,
    database = "neo4j",
)



```

### **Merge KEGG edges into neo4j DB**  
  
```{r merge kegg edges into neo4j DB}

path_to_edges <- "file:///Network_reasoning/variant_edgelist_intogen.csv"

## if edges don't exist, create them... 
cypher_MergeVariantEdges <- paste0("LOAD CSV WITH HEADERS FROM '",path_to_edges,"' AS edges
                                 MATCH (n1) WHERE edges.from = n1.label
                                 MATCH (n2:gene) WHERE edges.to = n2.entrez_symbol
                                 CALL {WITH n1, n2, edges
                                 CALL apoc.merge.relationship(n1,edges.relationship,
                                {LINK_FUNCTION:edges.LINK_FUNCTION}, 
                                {driver_mechanism:edges.driver_mechanism, LINK_FUNCTION:edges.LINK_FUNCTION, source:edges.source, refresh_date:edges.refresh_date}, 
                                n2, 
                                {source:edges.source, refresh_date:edges.refresh_date}
                                )
YIELD rel
RETURN NULL}
RETURN null")

## run query
neo4j_query(
    con = list(address = base_url, uid = username, pwd = password),
    qry = cypher_MergeVariantEdges,
    shell_path = shell_path,
    database = "neo4j",
)


```

   
`r knitr::knit_exit()`  





      
  